Notes on Prediction Types

### 2011-06-17 docs/pert_notes/pert.tex
fix multiply defined labels
fix section heading
fix EOS edge states
minor updates

@@

The full density evolution equation is
\begin{eqnarray}
\frac{\partial\rho}{\partial t} &=& -\nabla\cdot(\rho\Ub) \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub \, . \label{rho equation}
\end{eqnarray}
The species are evolved according to
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho\Ub X_k) + \rho \omegadot_k \nonumber \\
&=& -\Ub\cdot\nabla(\rho X_k) - \rho X_k \nabla\cdot\Ub + \rho \omegadot_k \, . \label{species equation}
\end{eqnarray}
In practice, only the species evolution equation is evolved, and the
total density is set as
\begin{equation}
\rho = \sum_k (\rho X_k)
\end{equation}
To advance $(\rho X_k)$ to the next timestep, we need to predict a
time-centered, interface state.  Algebraically, there are multiple
paths to get this interface state---we can predict $(\rho X)$ to the
edges as a single quantity, or predict $\rho$ and $X$ separately
(either in full or perturbation form).  In the notes below, we use the
subscript `{\tt edge}' to indicate what quantity was predicted to the
edges.  In MAESTRO, the different methods of computing $(\rho X)$ on
edges are controlled by the {\tt species\_pred\_type} parameter.  The
resulting edge state is shown in the following table:

\begin{table}[h]
\centering
\begin{tabular}{l|c}
\hline
\hline
{\tt species\_pred\_type} &   {$(\rho X_k)$ edge state} \\
\hline \\[-3mm]
1 / {\tt predict\_rhoprime\_and\_X}  &  $\left(\rho_0^{n+\myhalf,{\rm avg}}
+ \rho'_\mathrm{edge} \right)(X_k)_\mathrm{edge}$ \\[2mm]
2 / {\tt predict\_rhoX}  &  $(\rho X_k)_\mathrm{edge}$ \\[2mm]
3 / {\tt predict\_rho\_and\_X}  &  $\rho_\mathrm{edge} (X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table}
We note the labels {\tt predict\_rhoprime\_and\_X}, {\tt predict\_rhoX}, and
{\tt predict\_rho\_and\_X} are provided by the {\tt pred\_parameters}
module.

\subsection{Advective force for interface prediction}
In predicting the interface states, we work with an advection-like
equation, and call anything other than the advective term a ``force''.
For example, for a scalar $\phi$, obeying a conservation law with a source term, $f$:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\nabla \cdot (\phi \Ub) + f
\end{equation}
We write this in advective form as:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\Ub \cdot \nabla \phi \, \underbrace{- \phi \nabla \cdot \Ub}_{\text{``advective force''}} + f
\end{equation}
Here we identify the non-advection portion of the divergence of the
flux as the ``advective force'' for the
advective equation.  In predicting the interface states, this advective
force will be used in computing the interface states.  Depending on the
nature of $f$, it may either appear in the prediction as well or be absent
if the process it describes is incorporated by Strang-splitting.

@@

We predict both $\rho'$ and $\rho_0$ to edges separately and later use them to 
reconstruct $\rho$ at edges.  The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{``\rho_0 ~ \text{force"}}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{\rho' ~ \text{force}},
\label{rhoprime equation}
\end{equation}
where the underbraced ``$\rho'$ force'' is computed in {\tt modify\_scal\_force}.

@@

\subsubsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^{(1)} - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt density\_advance} and the
underbraced term in (\ref{rhoprime equation}) as the forcing.

@@

\item if we predicted $T$, convert this predicted
  edge state to an' intermediate ``enthalpy'' state (the quotes
  indicate that it may be perturbational or full enthalpy) by calling
  the EOS.

@@

\begin{landscape}
\begin{table}[t]
\caption{Summary of enthalpy edge state construction}
\label{table:pred:hoverview}
\renewcommand{\arraystretch}{1.5}
{\small
\centering
\begin{tabular}{l|l|c|c|c|c}
\hline
\hline
{\tt species\_pred\_type} &   
{\tt enthalpy\_pred\_type} &
{quantity predicted} &
{intermediate} &
{species quantity} &
{final $(\rho h)$} \\[-5pt]
 &
 &
{in {\tt make\_edge\_scal}} &
{``enthalpy'' edge state} &
{available in {\tt mkflux}} &
{edge state} \\
\hline 
1 / {\tt predict\_rhoprime\_and\_X}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\hline
2 / {\tt predict\_rhoX}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
2 / {\tt predict\_rhoX}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\end{tabular}
} % end \small
\end{table}
\end{landscape}

@@

\begin{table*}[h]
\centering
\caption{EOS states in {\tt makeHfromRhoT\_edge}\newline}
\label{table:pred:hforce}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c|c}
\hline
\hline
{\tt species\_pred\_type} & {$\rho$ edge state} &   {$X$ edge state}  \\
\hline
{\tt predict\_rhoprime\_and\_X} & 
  $\rho_0^{n+\myhalf,\rm{avg}} + \rho'_\mathrm{edge}$ & 
  $X_\mathrm{edge}$ \\
{\tt predict\_rhoX} & 
  $\rho_\mathrm{edge}$ & 
  $(\rho X)_\mathrm{edge}/\rho_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table*}

@@

>>>>

\section{General form of an advection equation}
In predicting the interface states, we work with an advection-like
equation, and call anything other than the advective term a ``force''.
For example, for a scalar $\phi$, obeying a conservation law with a source term, $f$:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\nabla \cdot (\phi \Ub) + f
\end{equation}
We write this in advective form as:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\Ub \cdot \nabla \phi \, \underbrace{- \phi \nabla \cdot \Ub}_{\text{``advective force''}} + f
\end{equation}
Here we identify the non-advection portion of the divergence of the
flux as the ``advective force'' for the advective equation.  This
advective force will be used as a source when predicting the edge
states.  Depending on the nature of $f$, it may either appear in the
prediction as well or be absent if the process it describes is
incorporated by Strang-splitting.

@@

The full density evolution equation is
\begin{eqnarray}
\frac{\partial\rho}{\partial t} &=& -\nabla\cdot(\rho\Ub) \nonumber \\
&=& -\Ub\cdot\nabla\rho - \rho\nabla\cdot\Ub \, . \label{rho equation}
\end{eqnarray}
The species are evolved according to
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho\Ub X_k) + \rho \omegadot_k \nonumber \\
&=& -\Ub\cdot\nabla(\rho X_k) - \rho X_k \nabla\cdot\Ub + \rho \omegadot_k \, . \label{species equation}
\end{eqnarray}
In practice, only the species evolution equation is evolved, and the
total density is set as
\begin{equation}
\rho = \sum_k (\rho X_k)
\end{equation}
To advance $(\rho X_k)$ to the next timestep, we need to predict a
time-centered, interface state.  Algebraically, there are multiple
paths to get this interface state---we can predict $(\rho X)$ to the
edges as a single quantity, or predict $\rho$ and $X$ separately
(either in full or perturbation form).  In the notes below, we use the
subscript `{\tt edge}' to indicate what quantity was {\em predicted} to the
edges.  In MAESTRO, the different methods of computing $(\rho X)$ on
edges are controlled by the {\tt species\_pred\_type} parameter.  The
quantities predicted to edges and the 
resulting edge state are shown in the table~\ref{table:pred:species}

\begin{table}[h]
\centering
\caption{Summary of species edge state construction}
\label{table:pred:species}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c|c}
\hline
\hline
{\tt species\_pred\_type} &   {quantities predicted} & {$(\rho X_k)$ edge state} \\[-5pt]
 & {in {\tt make\_edge\_scal}} & \\
\hline 
1 / {\tt predict\_rhoprime\_and\_X}  &  
  $\rho'_\mathrm{edge}$, $(X_k)_\mathrm{edge}$ &
  $\left(\rho_0^{n+\myhalf,{\rm avg}}  
  + \rho'_\mathrm{edge} \right)(X_k)_\mathrm{edge}$ \\
2 / {\tt predict\_rhoX}  &  
  $\rho'_\mathrm{edge}$, $(\rho X_k)_\mathrm{edge}$ &
  $(\rho X_k)_\mathrm{edge}$ \\
3 / {\tt predict\_rho\_and\_X}  &  
  $\rho_\mathrm{edge}$, $(X_k)_\mathrm{edge}$ &
  $\rho_\mathrm{edge} (X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table}
We note the labels {\tt predict\_rhoprime\_and\_X}, {\tt predict\_rhoX}, and
{\tt predict\_rho\_and\_X} are provided by the {\tt pred\_parameters}
module.

@@

We predict both $\rho'$ and $\rho_0$ to edges separately and later use them to 
reconstruct $\rho$ at edges.  The base state density evolution equation is
\begin{equation}
\frac{\partial\rho_0}{\partial t} = -\nabla\cdot(\rho_0 w_0 \eb_r) = 
-w_0\frac{\partial\rho_0}{\partial r} 
\underbrace{-\rho_0\frac{\partial w_0}{\partial r}}_{``\rho_0 ~ \text{force"}}.
\label{rho0 equation}
\end{equation}
Subtract (\ref{rho0 equation}) from (\ref{rho equation}) and rearrange
terms, noting that $\Ub = \Ubt + w_o\eb_r$, to obtain the
perturbational density equation,
\begin{equation}
\frac{\partial\rho'}{\partial t} = -\Ub\cdot\nabla\rho' \underbrace{- \rho'\nabla\cdot\Ub 
- \nabla\cdot(\rho_0\Ubt)}_{\rho' ~ \text{force}} \, .
\label{rhoprime equation}
\end{equation}

@@

\subsubsection{Predicting $\rho'$ at edges}
We define $\rho' = \rho^{(1)} - \rho_0^n$.  Then we predict $\rho'$ to
edges using {\tt make\_edge\_scal} in {\tt density\_advance} and the
underbraced term in Eq.~\ref{rhoprime equation} as the forcing.  This
force is computed in {\tt modify\_scal\_force}.

@@

\item if we predicted $T$, convert this predicted
  edge state to an intermediate ``enthalpy'' state (the quotes
  indicate that it may be perturbational or full enthalpy) by calling
  the EOS.

@@

\begin{landscape}
\begin{table}[t]
\caption{Summary of enthalpy edge state construction}
\label{table:pred:hoverview}
\renewcommand{\arraystretch}{1.5}
{\small
\centering
\begin{tabular}{l|l|c|c|c|c}
\hline
\hline
{\tt species\_pred\_type} &   
{\tt enthalpy\_pred\_type} &
{cell-centered} &
{intermediate} &
{species quantity} &
{final $(\rho h)$} \\[-5pt]
 &
 &
{quantity predicted} &
{``enthalpy''} &
{available in {\tt mkflux}} &
{edge state} \\[-5pt]
 &
 &
{in {\tt make\_edge\_scal}} &
{edge state} &
 &
 \\
\hline 
1 / {\tt predict\_rhoprime\_and\_X}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'_\mathrm{edge}$ & 
  $X_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left [ (\rho h)_0^{n+\myhalf,{\rm avg}} + (\rho h)'_\mathrm{edge} \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 2 / {\tt predict\_h} &
  $h$ & $h_\mathrm{edge}$ & 
  $X_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left ( \rho_0^{n+\myhalf,{\rm avg}} + \rho'_\mathrm{edge} \right ) h_\mathrm{edge}$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'_\mathrm{edge}$ & 
  $X_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left [ (\rho h)_0^{n+\myhalf,{\rm avg}} + (\rho h)'_\mathrm{edge} \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h_\mathrm{edge}$ & 
  $X_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left ( \rho_0^{n+\myhalf,{\rm avg}} + \rho'_\mathrm{edge} \right ) h_\mathrm{edge}$ \\
\hline
\hline
2 / {\tt predict\_rhoX}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'_\mathrm{edge}$ & 
  $(\rho X)_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left [ (\rho h)_0^{n+\myhalf,{\rm avg}} + (\rho h)'_\mathrm{edge} \right ]$ \\
2 / {\tt predict\_rhoX}  & 2 / {\tt predict\_h} &
  $h$ & $h_\mathrm{edge}$ & 
  $(\rho X)_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left ( \rho_0^{n+\myhalf,{\rm avg}} + \rho'_\mathrm{edge} \right ) h_\mathrm{edge}$ \\
2 / {\tt predict\_rhoX}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'_\mathrm{edge}$ & 
  $(\rho X)_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left [ (\rho h)_0^{n+\myhalf,{\rm avg}} + (\rho h)'_\mathrm{edge} \right ]$ \\
2 / {\tt predict\_rhoX}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h_\mathrm{edge}$ & 
  $(\rho X)_\mathrm{edge}$, $\rho'_\mathrm{edge}$ & 
  $\left ( \rho_0^{n+\myhalf,{\rm avg}} + \rho'_\mathrm{edge} \right ) h_\mathrm{edge}$ \\
\hline
\end{tabular}
} % end \small
\end{table}
\end{landscape}

@@

\begin{table*}[h]
\centering
\caption{EOS states in {\tt makeHfromRhoT\_edge}\newline}
\label{table:pred:EOSinputs}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c|c}
\hline
\hline
{\tt species\_pred\_type} & {$\rho$ edge state} &   {$X_k$ edge state}  \\
\hline
{\tt predict\_rhoprime\_and\_X} & 
  $\rho_0^{n+\myhalf,\rm{avg}} + \rho'_\mathrm{edge}$ & 
  $(X_k)_\mathrm{edge}$ \\
{\tt predict\_rhoX} & 
  $\sum_k (\rho X_k)_\mathrm{edge}$ & 
  $(\rho X_k)_\mathrm{edge}/\sum_k (\rho X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table*}

@@

\subsubsection{Converting $T_\mathrm{edge}$ to $h_\mathrm{edge}$}
